OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-03-19 paper/paper.Rmd
Update algorithm description

- **TODO: ask Sarah Westcott to check the accuracy of this description**.
- **TODO: toy example like opticlust with figure**.

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering.
**TODO: more narrative, less documentation-like.**
**TODO: extension of opticlust except ref otus are fixed. just like opticlust only query seqs can move otus.**
OptiFit takes as input a list of reference OTUs and their sequences, a list of
query sequences to assign to the reference OTUs, the sequence pairs that are
within the distance threshold (e.g. 0.03), and the metric to assess clustering
quality (default: MCC).
Query sequences are randomly seeded in reference OTUs, then for each sequence
the algorithm calculates the quality metric based on whether the sequence stays
in its current OTU, moves to each of the other OTUs, or is discarded.
If two or more OTU assignments are of equal quality, a random number generator
is used to break the tie.
This process is repeated until the quality metric stabilizes, changing by no
more than 0.0001 by default, or until a maximum number of iterations is reached
(default: 100).
In closed-reference mode, any query sequences that cannot be assigned are thrown
out (**TODO: exactly what determines whether a seq can't be assigned?**), and
the results only contain OTUs that exist in the original reference.
In open-reference mode, unassigned query sequences are clustered _de novo_ using
OptiClust to generate additional OTUs. The final quality score is reported with
the best OTU assignments.
There are two strategies for generating OTUs with OptiFit: 1) fit sequences to
reference OTUs of an independent database, or 2) split the dataset into a
reference and query fraction, cluster the reference sequences _de novo_, then
fit the query sequences to the reference OTUs.

@@

- TODO; start with problem with current state of the art. 
- tied to reference dbs taht may be unrelated to samples studying. can we use our own data as a reference?
- lay out the use case and how this is a benefit over ref-db clustering.
- we wanted ot use x% of data as ref, weren't sure what size to use and strategy for selecting.
- important result: it works!

@@

TODO; ref weight simple only here.
OTU quality from the split dataset strategy with OptiFit was highly similar to
that from _de novo_ clustering the whole dataset with OptiClust regardless of
mode (`r mcc_opticlust_vs_fit_split`% difference in median MCC).
OTU quality was remarkably stable across reference fraction sizes within
parameter sets (Fig. **XX**).
For example, splitting the human dataset 100 times yielded a coefficient of
variation of `r cv_fit_split_mcc_human` for the MCC score.
In terms of runtime, closed and open reference OptiFit performed faster than
OptiClust on whole datasets by `r sec_closed_fit_split_vs_clust`% and 
`r sec_open_fit_split_vs_clust` respectively.
The split dataset strategy performed `r sec_closed_fit_split_vs_db`% faster than
the database strategy in closed reference mode and `r sec_open_fit_split_vs_db`%
in open reference mode.
TODO: fraction mapped.

TODO: pick a fraction (e.g. 50%). this comparison would be supplemental if there's a figure.
We also tested three methods for selecting the fraction of sequences to be used
as the reference; a simple random sample, weighting sequences by relative
abundance, and weighting by similarity to other sequences in the dataset.
OTU quality was similar with the simple and abundance-weighted sampling (median
MCCs `r mcc_fit_split_simple` and `r mcc_fit_split_abun` respectively), but
worse for similarity-weighted sampling with a median MCC of `r mcc_fit_split_dist`.
In closed reference mode, the fraction of query sequences that can be fit to the
reference OTUs increases as the reference fraction increases; from 
`r frac_fit_split_0.1`% of query sequences fit when using 10% of the dataset as 
the reference, to `r frac_fit_split_0.8`% of query sequences fit when using 80% 
of the dataset as the reference.

>>>>

- **TODO: ask Sarah Westcott to check the accuracy of this description**.

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering.
OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
their sequences, a list of query sequences to assign to the reference OTUs, and
the sequence pairs that are within the distance threshold (e.g. 0.03).
Query sequences are randomly seeded in reference OTUs, then for each sequence
the algorithm calculates the quality metric (default: MCC) based on whether the
sequence stays in its current OTU, moves to each of the other OTUs, or is
discarded.
If two or more OTU assignments are of equal quality, a random number generator
is used to break the tie.
This process is repeated until the quality metric stabilizes, changing by no
more than 0.0001 by default, or until a maximum number of iterations is reached
(default: 100).
In closed-reference mode, any query sequences that cannot be assigned are thrown
out (**TODO: exactly what determines whether a seq can't be assigned?**), and
the results only contain OTUs that exist in the original reference.
In open-reference mode, unassigned query sequences are clustered _de novo_ using
OptiClust to generate additional OTUs. The final quality score is reported with
the best OTU assignments.
There are two strategies for generating OTUs with OptiFit: 1) fit sequences to
reference OTUs of an independent database, or 2) split the dataset into a
reference and query fraction, cluster the reference sequences _de novo_, then
fit the query sequences to the reference OTUs.

@@

- **TODO**; start with problem with current state of the art. 
- tied to reference dbs taht may be unrelated to samples studying. can we use our own data as a reference?
- lay out the use case and how this is a benefit over ref-db clustering.
- we wanted ot use x% of data as ref, weren't sure what size to use and strategy for selecting.
- important result: it works!

@@

**TODO; ref weight simple only here.**
OTU quality from the split dataset strategy with OptiFit was highly similar to
that from _de novo_ clustering the whole dataset with OptiClust regardless of
mode (`r mcc_opticlust_vs_fit_split`% difference in median MCC).
OTU quality was remarkably stable across reference fraction sizes within
parameter sets (Fig. **XX**).
For example, splitting the human dataset 100 times yielded a coefficient of
variation of `r cv_fit_split_mcc_human` for the MCC score.
In terms of runtime, closed and open reference OptiFit performed faster than
OptiClust on whole datasets by `r sec_closed_fit_split_vs_clust`% and 
`r sec_open_fit_split_vs_clust` respectively.
The split dataset strategy performed `r sec_closed_fit_split_vs_db`% faster than
the database strategy in closed reference mode and `r sec_open_fit_split_vs_db`%
in open reference mode.
TODO: fraction mapped.

**TODO: pick a fraction (e.g. 50%). this part is less important. figure would be supplemental.**
We also tested three methods for selecting the fraction of sequences to be used
as the reference; a simple random sample, weighting sequences by relative
abundance, and weighting by similarity to other sequences in the dataset.
OTU quality was similar with the simple and abundance-weighted sampling (median
MCCs `r mcc_fit_split_simple` and `r mcc_fit_split_abun` respectively), but
worse for similarity-weighted sampling with a median MCC of `r mcc_fit_split_dist`.
In closed reference mode, the fraction of query sequences that can be fit to the
reference OTUs increases as the reference fraction increases; from 
`r frac_fit_split_0.1`% of query sequences fit when using 10% of the dataset as 
the reference, to `r frac_fit_split_0.8`% of query sequences fit when using 80% 
of the dataset as the reference.
